---
title: "LASSO Regression"
output: html_notebook
---

LASSO regression using thresholded NMF corr genes? 

## Regress on NMF score

Pseudobulk Expression

```{r}
pseudobulk_vst <- readRDS("../BALLbulk_Deconvolution/BALL_89pt_pseudobulk_vst.rds")
pseudobulk_vst
```

```{r}
read_csv('../CompositionAnalysis/BALL_Composition_DevState_NMFscores.csv') 
```


```{r}
read_csv('../CompositionAnalysis/BALL_Composition_DevState_NMFscores.csv') %>% 
  select(TB, DevState__HSC_MPP = NMF8_HSCMPPLMPP, DevState__Myeloid_Prog = NMF6_MyeloidProg, DevState__Pre_pDC = NMF7_pDC, 
         DevState__Early_Lymphoid = NMF2_EarlyLymphoid, DevState__Pro_B = NMF1_ProB, DevState__Pre_B = NMF4_PreB) %>% 
  write_csv('../BALLmap_paper/SupTables/ST12_scBALL_DevState_NMF_Abundance_Scores.csv')
```


```{r}
NMF_ptscores <- read_csv('../CompositionAnalysis/BALL_Composition_DevState_NMFscores.csv') %>% 
  left_join(read_csv('scBALL_IDconversion.csv')) %>%
  select(ID_Bulk = ID, contains('NMF')) %>%
  pivot_longer(-ID_Bulk, names_to = 'NMF', values_to = 'NMFscore') %>% 
  left_join(pseudobulk_vst@meta.data %>% select(ID_Bulk) %>% rownames_to_column('ID_scRNA') ) %>% 
  mutate(Lineage = NMF %>% str_replace('.*_',''), NMF = NMF %>% str_replace('_.*','')) %>% 
  select(Patient = ID_scRNA, NMF, Lineage, NMFscore) 

NMF_ptscores
```

```{r}
NMFcorr <- read_csv('NMF_GeneCorr_Thresholding.csv') %>% 
  filter(threshold == 'pass', qvalue < 0.01) %>% arrange(qvalue) %>% arrange(NMF)
NMFcorr
```

### Nested Cross Validation by Feature set

Define feature space - load markers 


```{r}
LinDE_FDR01_genes <- read_csv('BALL_DEresults_NMF_Lineage.csv') %>% filter(padj < 0.01, stat > 0) %>% pull(Gene) %>% unique()
LinDE_FDR05_genes <- read_csv('BALL_DEresults_NMF_Lineage.csv') %>% filter(padj < 0.05, stat > 0) %>% pull(Gene) %>% unique()
BDevDE_FDR01_genes <- read_csv('../BDevelopment_Characterization/BDevelopment_CellType_DEresults.csv') %>% 
  filter(padj < 0.01, stat > 0) %>% pull(Gene) %>% unique()
BDevDE_FDR05_genes <- read_csv('../BDevelopment_Characterization/BDevelopment_CellType_DEresults.csv') %>% 
  filter(padj < 0.05, stat > 0) %>% pull(Gene) %>% unique()
```


```{r}
read_csv('../BDevelopment_Characterization/BDevelopment_CellType_DEresults.csv') %>% 
  filter(padj < 0.01, stat > 0) %>% pull(Gene) %>% unique() %>% length()
```


```{r}
NMFcorr <- NMFcorr %>% mutate(LinDE_FDR05 = Gene %in% LinDE_FDR05_genes, 
                              LinDE_FDR01 = Gene %in% LinDE_FDR01_genes, 
                              BDevDE_FDR01 = Gene %in% BDevDE_FDR01_genes, 
                              BDevDE_FDR05 = Gene %in% BDevDE_FDR05_genes)
NMFcorr
```

```{r}
NMFcorr %>% filter(LinDE_FDR01==TRUE, BDevDE_FDR01==TRUE) %>% select(Gene) %>% unique()
```


```{r}
evaluate_model <- function(model, x_val, anno_val, lambda, 
                           feature_name, iteration, foldname){

  # Create score classification with survival and get covariates
  pred_y <- predict(model, x_val, s = lambda) %>% data.frame()
  colnames(pred_y) <- 'PredScore'
  pred_y <- pred_y %>% rownames_to_column('Patient') %>% 
    # add anno to get covariates
    left_join(anno_val, by = 'Patient')
  
  # Calculate correlation in validation set
  pearson <- cor(pred_y$PredScore, pred_y$NMFscore, method = 'pearson')
  spearman <- cor(pred_y$PredScore, pred_y$NMFscore, method = 'spearman')
  
  # Summary Metrics
  summary_metrics <- data.frame(
    'model_id' = paste0(feature_name, '_iter', iteration, '_', foldname),
    'lambda' = lambda,
    'model_size' = sum(coef(model, s = lambda)!=0),
    'pearson' = pearson,
    'spearman' = spearman,
    'features' = feature_name,
    'iteration' = iteration,
    'foldname' = foldname
  )
  return(summary_metrics)
}

```


```{r}
gridsearch_lasso <- function(expr_train, expr_val, anno_train, anno_val, features, feature_name,
                             iteration, foldname, summary_metrics){
  
  # Filter expr matrix for feature set
  x_train <- expr_train[, colnames(expr_train) %in% features]
  x_val <- expr_val[, colnames(expr_val) %in% features]

  # Train LASSO 
  model <- train_LASSO(x_train, anno_train)

  # Get summary metrics for lambda.min and lambda.1se
  for(lambda in c('lambda.min', 'lambda.1se')){
    summary_metrics <- summary_metrics %>% rbind(
      evaluate_model(model = model, x_val = x_val, anno_val = anno_val, lambda = lambda, 
                     feature_name = feature_name, iteration = iteration, foldname = foldname))
  }
  
  return(summary_metrics)
}

```


```{r}
nestedCV_regression <- function(train_anno, train_expr, iteration, feature_sets, summary_metrics){
  # set up random seed and shuffle data 
  set.seed(iteration)
  train_anno <- train_anno[sample(nrow(train_anno)),]
  train_expr <- train_expr[sample(nrow(train_expr)),]
  
  ## 5-fold outer cross validation
  folds <- rsample::vfold_cv(train_anno, 5)
  for(outer_cv in 1:5){
    # fold ID
    foldname <- folds$id[[outer_cv]]
    # get anno splits
    anno_train <- analysis(folds$splits[[outer_cv]])
    anno_val <- assessment(folds$splits[[outer_cv]])
    # get expr splits
    expr_train <- train_expr[anno_train$Patient,]
    expr_val <- train_expr[anno_val$Patient,]
    
    # Iterate through feature set and run gridsearch to train survival functions
    for(feature_name in names(feature_sets)){
      # get feature list
      features <- feature_sets[[feature_name]]
      # run gridsearch and get results
      summary_metrics <- gridsearch_lasso(expr_train = expr_train, expr_val = expr_val, anno_train = anno_train, anno_val = anno_val, 
                                  features = features, feature_name = feature_name, iteration = iteration, foldname = foldname,
                                  summary_metrics = summary_metrics)
    }
  }
  return(summary_metrics)
}

```


```{r}
library(tidymodels)
library(glmnet)

output <- data.frame()
train_x <- pseudobulk_vst@assays$RNA@data[,unique(NMF_ptscores$Patient)] %>% data.matrix() %>% t()

for(NMFcomp in c('NMF1', 'NMF2', 'NMF3', 'NMF4', 'NMF5', 'NMF6', 'NMF7', 'NMF8', 'NMF9', 'NMF10')){
  
  print(paste0('NMF Component: ', NMFcomp))
  
  temp_output <- data.frame()
  
  train_y <- NMF_ptscores %>% filter(NMF == NMFcomp) %>% select(Patient, NMFscore)
  featurespace <- list('PosCorr_LinDE_FDR05' = NMFcorr %>% filter(NMF == NMFcomp, LinDE_FDR05 == TRUE, pearson > 0) %>% pull(Gene), 
                       'PosCorr_LinDE_FDR01' = NMFcorr %>% filter(NMF == NMFcomp, LinDE_FDR01 == TRUE, pearson > 0) %>% pull(Gene), 
                       'PosCorr_LinDE_BDevDE_FDR05' = NMFcorr %>% filter(NMF == NMFcomp, LinDE_FDR05 == TRUE, BDevDE_FDR05 == TRUE, pearson > 0) %>% pull(Gene), 
                       'PosCorr_LinDE_BDevDE_FDR01' = NMFcorr %>% filter(NMF == NMFcomp, LinDE_FDR01 == TRUE, BDevDE_FDR01 == TRUE, pearson > 0) %>% pull(Gene), 
                       'AnyCorr_LinDE_FDR05' = NMFcorr %>% filter(NMF == NMFcomp, LinDE_FDR05 == TRUE) %>% pull(Gene),
                       'AnyCorr_LinDE_FDR01' = NMFcorr %>% filter(NMF == NMFcomp, LinDE_FDR01 == TRUE) %>% pull(Gene),
                       'AnyCorr_LinDE_BDevDE_FDR05' = NMFcorr %>% filter(NMF == NMFcomp, LinDE_FDR05 == TRUE, BDevDE_FDR05 == TRUE) %>% pull(Gene), 
                       'AnyCorr_LinDE_BDevDE_FDR01' = NMFcorr %>% filter(NMF == NMFcomp, LinDE_FDR01 == TRUE, BDevDE_FDR01 == TRUE) %>% pull(Gene)
  )
  
  for(iteration in 1:10){
    print(paste0('iteration ', iteration))
    temp_output <- nestedCV_regression(train_anno = train_y, train_expr = train_x, iteration = iteration, feature_sets = featurespace, 
                                       summary_metrics = temp_output) 
  }
  ## annotate and add to final output
  output <- bind_rows(output, temp_output %>% mutate(NMF = NMFcomp))
}

output %>% write_csv('RepNestedCV_results_NMFregression.csv')
```


**5-Fold cross validation with 10 repeats within the pseudobulk to estimate the best parameters and get a gestalt of the overall accuracy**
After choosing the best combination of parameters we will test on the bulk RNA-seq dataset.


```{r}
output <- read_csv('RepNestedCV_results_NMFregression.csv') 
output %>% pull(features) %>% table()
```

```{r, fig.height = 6, fig.width = 12}
output %>% 
  mutate(NMF = factor(NMF, levels = c('NMF1', 'NMF2', 'NMF3', 'NMF4', 'NMF5', 'NMF6', 'NMF7', 'NMF8', 'NMF9', 'NMF10'))) %>% 
  ggplot(aes(x = reorder(features, -model_size), y = model_size, fill = lambda)) + 
  geom_hline(yintercept = 10, lty = 2) + geom_hline(yintercept = 20, lty = 2) + 
  geom_hline(yintercept = 30, lty = 2) + geom_hline(yintercept = 40, lty = 2) + 
  geom_boxplot(outlier.size = 0.8) + ggbeeswarm::geom_quasirandom(dodge.width = 0.7, size = 0.2, alpha = 0.7) + 
  facet_wrap(.~NMF, ncol = 6) + theme_pubr() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + 
  stat_compare_means(label = 'p.signif')
  
```


```{r, fig.height = 6, fig.width = 12}
output %>% 
  mutate(NMF = factor(NMF, levels = c('NMF1', 'NMF2', 'NMF3', 'NMF4', 'NMF5', 'NMF6', 'NMF7', 'NMF8', 'NMF9', 'NMF10'))) %>% 
  ggplot(aes(x = reorder(features, -pearson), y = pearson, fill = lambda)) + 
  geom_boxplot(outlier.size = 0.8) + ggbeeswarm::geom_quasirandom(dodge.width = 0.7, size = 0.2, alpha = 0.7) + 
  facet_wrap(.~NMF, ncol = 6) + theme_pubr() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + 
  stat_compare_means(label = 'p.signif')
  
```


```{r}
NMFconvert
```


In summary, using combination of positively and negatively correlated genes leads to better performance, particularly for: 
  NMF1 (Pro-B)
  NMF5 (Erythroid)
  NMF11 (Naive T)
  
In terms of filtering of correlated genes; Lin DE BDev DE FDR < 0.01 as a filter consistently lead to the best performance. 

Lambda choice of Minimum + 1SE resulted in model size reductions of nearly 20 genes without sacrificing performance by nested CV. 

```{r}
output_CVmedians <- output %>% filter(lambda == 'lambda.1se', features == 'AnyCorr_LinDE_BDevDE_FDR01') %>% #AnyCorr_LinDE_FDR01. AnyCorr_LinDE_BDevDE_FDR01
  select(NMF, model_size, pearson, spearman) %>% 
  group_by(NMF) %>% summarise_all(median) %>% arrange(pearson)

output_CVmedians 
```



### Final Choice: Corr Threshold + Lineage DE FDR01; lambda min + 1SE

Train from both positively corr and any corr. If models from both approaches have negative coefficients, use any corr. 
Also make sure that the genes are present in the bulk RNAseq data


```{r}
train_LASSO <- function(x_train, y_train, alpha = 1){
  
  train_y <- y_train$NMFscore
  
  # Perform Lasso regression with LOOCV 
  model <- cv.glmnet(x = x_train, y = train_y, nfold = dim(x_train)[1], family = 'gaussian', alpha = alpha, maxit=1000000, standardize=FALSE)
  #plot(model)

  return(model)
}
```

```{r}
bulkRNAgenes <- data.table::fread("../BALLbulk_Deconvolution/BALL_bulkRNA_data/BALL_BulkRNAseq_subsetgene_rawcounts.txt")$Gene
bulkRNAgenes %>% length()
```


```{r}
set.seed(123)
model_list <- list()
# subset NMF corr with genes present in the bulk RNA data
NMFcorr <- NMFcorr %>% filter(Gene %in% bulkRNAgenes)

for(NMFcomp in c('NMF1', 'NMF2', 'NMF3', 'NMF4', 'NMF5', 'NMF6', 'NMF7', 'NMF8', 'NMF9', 'NMF10')){
  
  # Define y variable; NMF score
  train_y <- NMF_ptscores %>% filter(NMF == NMFcomp) %>% select(Patient, NMFscore)
  # Define feature space to train from
  feature_space <- NMFcorr %>% filter(NMF == NMFcomp, LinDE_FDR01 == TRUE, BDevDE_FDR01 == TRUE) %>% pull(Gene) #BDevDE_FDR01 == TRUE
  # subset training set
  train_x <- pseudobulk_vst@assays$RNA@data[feature_space, train_y$Patient] %>% data.matrix() %>% t()
  
  model <- train_LASSO(train_x, y_train = train_y)
  model_list[[NMFcomp]] <- model
}
```

```{r}
# model weights 
modelweights <- data.frame()

for(NMFcomp in c('NMF1', 'NMF2', 'NMF3', 'NMF4', 'NMF5', 'NMF6', 'NMF7', 'NMF8', 'NMF9', 'NMF10')){
  modelweights <- modelweights %>% bind_rows(
    model_list[[NMFcomp]] %>% coef(s = 'lambda.1se') %>% data.matrix() %>% 
      data.frame() %>% dplyr::rename(Weight = s1) %>% rownames_to_column('Gene') %>% 
      tail(-1) %>% filter(Weight != 0) %>% arrange(-Weight) %>% mutate(Model = NMFcomp)
  )
}

modelweights <- modelweights %>% select(Model, Gene, Weight)
modelweights %>% group_by(Model) %>% summarise(count = n())
```


```{r}
# model weights 
modelweights <- data.frame()

for(NMFcomp in c('NMF1', 'NMF2', 'NMF3', 'NMF4', 'NMF5', 'NMF6', 'NMF7', 'NMF8', 'NMF9', 'NMF10')){
  modelweights <- modelweights %>% bind_rows(
    model_list[[NMFcomp]] %>% coef(s = 'lambda.1se') %>% data.matrix() %>% 
      data.frame() %>% dplyr::rename(Weight = s1) %>% rownames_to_column('Gene') %>% 
      tail(-1) %>% filter(Weight != 0) %>% arrange(-Weight) %>% mutate(Model = NMFcomp)
  )
}

modelweights <- modelweights %>% select(Model, Gene, Weight)
modelweights %>% group_by(Model) %>% summarise(count = n())
```


```{r}
modelweights %>% 
  left_join(NMFconvert %>% dplyr::rename(Model = NMF)) %>%
  mutate(coefficient = ifelse(Weight > 0, 'Positive', 'Negative') %>% factor(levels = c('Positive', 'Negative'))) %>% 
  group_by(NMFnamed, coefficient) %>% summarise(count = n()) %>%
  ggplot(aes(x = NMFnamed, y = count, fill = coefficient)) + geom_col() + ggpubr::theme_pubr() + 
  ggsci::scale_fill_simpsons() + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```





```{r}
NMFnamed_levels <- c('HSC_MPP', 'Myeloid_Prog', 'Pre_pDC', 'Early_Lymphoid', 'Pro_B', 'Pre_B', 
                      'Mature_B', 'Erythroid', 'Monocyte', 'T_NK')

NMFconvert <- data.frame(
  'NMF' = c('NMF8', 'NMF6', 'NMF7', 'NMF2', 'NMF1', 'NMF4', 
            'NMF5', 'NMF3', 'NMF9', 'NMF10') %>% factor(),
  'NMFnamed' = NMFnamed_levels %>% factor(levels = NMFnamed_levels)
)

NMFconvert
```

```{r}
modelweights <- modelweights %>% left_join(NMFconvert %>% dplyr::rename(Model = NMF)) %>% 
  arrange(Gene) %>% arrange(NMFnamed) %>% pivot_wider(id_cols=Gene, names_from=NMFnamed, values_from=Weight) %>% replace(is.na(.), 0) 

modelweights %>% write_csv("NMF_Lasso_ModelWeights.csv")
modelweights
```

## Figure out NMF scoring and validate on pseudobulk

```{r}
calculate_NMFscores = function(query, modelweights, scale = TRUE, sampleID = 'Patient'){
  
  # Check for overlap with model genes and query genes
  querygenes <- rownames(query)
  modelweights_missing <- sum(!(modelweights$Gene %in% querygenes))
  # check for missing genes
  if(modelweights_missing > 0){
    print(paste0('Warning: ', modelweights_missing, ' genes from NMF models are missing from query dataset'))
  }
  
  # filter model weights
  modelweights <- modelweights %>% filter(Gene %in% querygenes)
  modelweights_mat <- modelweights %>% column_to_rownames('Gene') %>% data.matrix()
  
  # multiply query by NMF lasso weights
  scored <- (t(query[modelweights$Gene,]) %*% modelweights_mat) %>% data.matrix() 
  if(scale == TRUE){
    scored <- scale(scored)
  }
  scored <- scored %>% as.data.frame() %>% rownames_to_column(sampleID) 
  
  return(scored)
}
```


```{r}
modelweights <- read_csv("NMF_Lasso_ModelWeights.csv")
modelweights
```

```{r}
pseudobulk_vst <- readRDS("../BALLbulk_Deconvolution/BALL_89pt_pseudobulk_vst.rds")
# calculate NMF scores from vst-normalized data
pseudobulk_vst_NMFscores <- calculate_NMFscores(pseudobulk_vst@assays$RNA@data, modelweights, scale = T, sampleID = 'Patient')
pseudobulk_vst_NMFscores[1:10,1:10]
```

**Training results - pseudobulk**

```{r, fig.height = 5, fig.width = 12}
NMF_compare <- NMF_ptscores %>% 
  left_join(NMFconvert) %>% 
  left_join(pseudobulk_vst_NMFscores %>% pivot_longer(-Patient, names_to = 'NMFnamed', values_to = 'predNMF')) 

NMF_compare %>% 
  mutate(NMFnamed = factor(NMFnamed, levels = NMFnamed_levels)) %>% 
  ggplot(aes(x = predNMF, y = NMFscore)) + 
  geom_point() + geom_smooth(method = 'lm') + 
  facet_wrap(.~NMFnamed, scales = 'free') + 
  theme_pubr() + stat_cor()
```




### True Bulk RNAseq "validation"

```{r}
bulkRNA_counts <- data.table::fread("../BALLbulk_Deconvolution/BALL_bulkRNA_data/BALL_BulkRNAseq_subsetgene_rawcounts.txt") %>% 
  column_to_rownames('Gene') %>% data.matrix()
bulkRNA_counts[1:10,1:10]
```

```{r}
library(DESeq2)
bulkRNA_dds <- DESeqDataSetFromMatrix(bulkRNA_counts[rowSums(bulkRNA_counts) >= 10,], 
                       colData = data.frame('Patient' = colnames(bulkRNA_counts)) %>% column_to_rownames('Patient'), 
                       design = ~1)
bulkRNA_vst <- assay(vst(bulkRNA_dds))
rm(bulkRNA_counts, bulkRNA_dds)

bulkRNA_vst[1:10,1:10]
```

```{r}
# calculate NMF scores from vst-normalized data
bulkRNA_vst_NMFscores <- calculate_NMFscores(bulkRNA_vst, modelweights, scale = T, sampleID = 'Patient')
bulkRNA_vst_NMFscores[1:10,1:10]
```

**Validation results - matched bulkRNAseq**

```{r}
NMF_ptscores_converted <- NMF_ptscores %>% 
  left_join(read_delim("../BALL_metadata_20230105.txt", delim = '\t') %>% select(Patient = Directory, Sample, ID, TB)) 

NMF_ptscores_compareBulkRNA <- NMF_ptscores_converted %>% 
  left_join(NMFconvert) %>% 
  left_join(bulkRNA_vst_NMFscores %>% pivot_longer(-Patient, names_to = 'NMFnamed', values_to = 'predNMF') %>% 
              dplyr::rename(ID = Patient)) 

NMF_ptscores_compareBulkRNA
```


```{r, fig.height = 3.5, fig.width = 12}
NMF_ptscores_compareBulkRNA %>% 
  mutate(NMFnamed = factor(NMFnamed, NMFnamed_levels)) %>% 
  ggplot(aes(x = predNMF, y = NMFscore)) + 
  geom_point() + geom_smooth(method = 'lm') + 
  facet_wrap(.~NMFnamed, scales = 'free', ncol = 5) + 
  theme_pubr() + stat_cor() + 
  ylab('NMF Lineage Score (scRNA composition)') + xlab('Predicted NMF Score (matched bulk RNA-seq)')
```



```{r, fig.height = 6, fig.width = 4}
NMF_ptscores_compareBulkRNA %>% 
  mutate(NMFnamed = factor(NMFnamed, NMFnamed_levels)) %>% 
  ggplot(aes(x = predNMF, y = NMFscore)) + 
  geom_point() + geom_smooth(method = 'lm') + 
  facet_wrap(.~NMFnamed, scales = 'free', ncol = 2) + 
  theme_pubr() + stat_cor() + 
  ylab('NMF Lineage Score (scRNA composition)') + xlab('Predicted NMF Score (matched bulk RNA-seq)')
```

Not bad!!
Save bulk RNAseq NMF scores: 

```{r}
bulkRNA_vst_NMFscores %>% write_csv('Bulk2046_NMFregression_LineageScores.csv')
bulkRNA_vst_NMFscores
```

```{r}
bulkRNA_vst_NMFscores %>% select(Patient, DevState__HSC_MPP = HSC_MPP, DevState__Myeloid_Prog = Myeloid_Prog, DevState__Pre_pDC = Pre_pDC, 
                                 DevState__Early_Lymphoid = Early_Lymphoid, DevState__Pro_B = Pro_B, DevState__Pre_B = Pre_B) %>% 
  write_csv('../BALLmap_paper/SupTables/ST15_BALL_Bulk2046_DevState_Abundance_Scores.csv')
```






```{r, fig.height = 3, fig.width = 7}
NMF_ptscores_compareBulkRNA %>% 
  group_by(Lineage) %>% 
  mutate(NMFscore_scaled = (NMFscore - mean(NMFscore)) / sd(NMFscore)) %>%
  filter(NMFnamed %in% c('HSC_MPP', 'Myeloid_Prog', 'Pre_pDC', 'Early_Lymphoid', 'Pro_B', 'Pre_B')) %>%
  mutate(NMFnamed = NMFnamed %>% str_replace('HSC_MPP', 'HSC / MPP') %>% str_replace('Early_','Early ') %>% str_replace('Myeloid_', 'Myeloid ') %>% 
           str_replace('_','-'), 
         NMFnamed = factor(NMFnamed, c('HSC / MPP', 'Myeloid Prog', 'Pre-pDC',  'Early Lymphoid',  'Pro-B', 'Pre-B'))) %>% 
  ggplot(aes(x = predNMF, y = NMFscore_scaled)) + 
  geom_point() + geom_smooth(method = 'lm') + 
  facet_wrap(.~NMFnamed, scales = 'free', ncol = 3) + 
  theme_pubr() + stat_cor() + theme(strip.text.x = element_text(size = 11.5)) + 
  ylab('NMF Lineage Score (scRNA composition)') + xlab('Predicted NMF Score (matched bulk RNA-seq)')

ggsave('BulkDecon_Figures/NMFregression_Benchmarking_bulkRNA_vs_scRNA_composition_wide.pdf', height = 4.8, width = 10.5, device = 'pdf')
```


```{r, fig.height = 5, fig.width = 3.5}
NMF_ptscores_compareBulkRNA %>% 
  filter(NMFnamed %in% c('HSC_MPP', 'Myeloid_Prog', 'Pre_pDC', 'Early_Lymphoid', 'Pro_B', 'Pre_B')) %>%
  mutate(NMFnamed = NMFnamed %>% str_replace('HSC_MPP', 'HSC / MPP') %>% str_replace('Early_','Early ') %>% str_replace('Myeloid_', 'Myeloid ') %>% 
           str_replace('_','-'), 
         NMFnamed = factor(NMFnamed, c('HSC / MPP', 'Early Lymphoid',  'Myeloid Prog', 'Pro-B', 'Pre-pDC', 'Pre-B'))) %>% 
  ggplot(aes(x = predNMF, y = NMFscore)) + 
  geom_point() + geom_smooth(method = 'lm') + 
  facet_wrap(.~NMFnamed, scales = 'free', ncol = 2) + 
  theme_pubr() + stat_cor() + 
  ylab('NMF Lineage Score (scRNA composition)') + xlab('Predicted NMF Score (matched bulk RNA-seq)')

ggsave('BulkDecon_Figures/NMFregression_Benchmarking_bulkRNA_vs_scRNA_composition.pdf', height = 8, width = 8, device = 'pdf')
```



```{r}
output %>% 
  left_join(NMFconvert) %>%
  filter(NMFnamed %in% c('HSC_MPP', 'Myeloid_Prog', 'Pre_pDC', 'Early_Lymphoid', 'Pro_B', 'Pre_B')) %>% 
  mutate(NMFnamed = NMFnamed %>% str_replace('HSC_MPP', 'HSC / MPP') %>% str_replace('Early_','Early ') %>% 
           str_replace('Myeloid_Prog', 'Early Myeloid') %>% str_replace('Pre_pDC', 'Plasmacytoid DC') %>% 
           str_replace('_','-'), 
         NMFnamed = factor(NMFnamed, c('HSC / MPP', 'Early Lymphoid',  'Myeloid Prog', 'Pro-B', 'Pre-pDC', 'Pre-B'))) %>% 
  filter(features == 'AnyCorr_LinDE_BDevDE_FDR01', lambda == 'lambda.1se') %>% 
  ggplot(aes(x = NMFnamed, y = pearson, fill = NMFnamed)) + 
  theme_pubr(legend = 'none') + geom_hline(yintercept = 0.7, lty = 2) + geom_hline(yintercept = 0.8, lty = 2) + geom_hline(yintercept = 0.9, lty = 2) + 
  geom_boxplot(outlier.size = 0) + ggbeeswarm::geom_quasirandom(width = 0.3, size = 0.8, alpha = 0.7) + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) + scale_y_continuous(breaks = c(0,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1)) + 
  xlab('Developmental State') + ylab('Pearson Correlation')
  
ggsave('BulkDecon_Figures/NMFregression_Benchmarking_CrossValidation_correlation.pdf', height = 3.8, width = 4.5, device='pdf')
```


```{r}
output %>% 
  left_join(NMFconvert) %>%
  #mutate(NMF = factor(NMF, levels = c('NMF1', 'NMF2', 'NMF3', 'NMF4', 'NMF5', 'NMF6', 'NMF7', 'NMF8', 'NMF9', 'NMF10'))) %>% 
  filter(features == 'AnyCorr_LinDE_BDevDE_FDR01', lambda == 'lambda.1se') %>% 
  ggplot(aes(x = NMFnamed, y = pearson, fill = NMFnamed)) + 
  theme_pubr(legend = 'none') + geom_hline(yintercept = 0.5, lty = 2) + geom_hline(yintercept = 0.75, lty = 2) + geom_hline(yintercept = 0.9, lty = 2) + 
  geom_boxplot(outlier.size = 0) + ggbeeswarm::geom_quasirandom(size = 1, alpha = 0.7) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) 
  
```


```{r}
output %>% 
  left_join(NMFconvert) %>%
  #mutate(NMF = factor(NMF, levels = c('NMF1', 'NMF2', 'NMF3', 'NMF4', 'NMF5', 'NMF6', 'NMF7', 'NMF8', 'NMF9', 'NMF10'))) %>% 
  filter(features == 'AnyCorr_LinDE_BDevDE_FDR01', lambda == 'lambda.1se') %>% 
  ggplot(aes(x = reorder(NMFnamed, -pearson), y = pearson, fill = NMFnamed)) + 
  theme_pubr(legend = 'none') + geom_hline(yintercept = 0.5, lty = 2) + geom_hline(yintercept = 0.75, lty = 2) + geom_hline(yintercept = 0.9, lty = 2) + 
  geom_boxplot(outlier.size = 0) + ggbeeswarm::geom_quasirandom(size = 1, alpha = 0.7) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) 
  
```

```{r}
NMF_ptscores_compareBulkRNA %>% drop_na() %>% 
  group_by()
```






## quickly evaluate on cord blood sorted

```{r}
cb_fractions = data.table::fread('../../../../../../../Dormancy/HSC_analysis/HSC_byOntogeny/Dicklab_sortedFractions_Batch4_CB_Hierarchy_vst.csv')
cb_fractions <- cb_fractions %>% column_to_rownames('Gene') %>% data.matrix()
cb_fractions %>% dim()
```


```{r}
CBfractions_vst_NMFscores <- calculate_NMFscores(cb_fractions, modelweights, scale = T, sampleID = 'Sample')
CBfractions_vst_NMFscores
```

```{r, fig.height = 8, fig.width = 10}
CBfractions_vst_NMFscores %>% 
  pivot_longer(-Sample, names_to = 'NMFsig', values_to = 'Score') %>% 
  mutate(NMFsig = factor(NMFsig, levels = NMFnamed_levels), 
         Population = Sample %>% str_replace('.*CB_',''),
         Population = factor(Population, levels = c('HSC', 'MPP', 'LMPP', 'CMP', 'GMP', 'MLPII', 'EarlyProB', 'PreProB', 
                                                    'ProB', 'PreB', 'B', #'B1', 'B2', 'B3', 'B4', 'B5', 'B6', 
                                                    'T', 'NK', 'EryP', 'Mono', 'Gr'))) %>% 
  filter(Population != 'NA') %>% 
  ggplot(aes(x = Population, y = Score, fill = Population)) + 
  geom_boxplot() + geom_jitter() +
  theme_pubr(legend = 'none') + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  facet_wrap(.~NMFsig, scales = 'free', ncol = 3)
  
```

```{r}
CBfractions_vst_NMFscores %>% 
  pivot_longer(-Sample, names_to = 'NMFsig', values_to = 'Score') %>% 
  mutate(NMFsig = factor(NMFsig, levels = NMFnamed_levels), 
         Population = Sample %>% str_replace('.*CB_',''),
         Population = factor(Population, levels = c('HSC', 'MPP', 'LMPP', 'CMP', 'GMP', 'MLPII', 'EarlyProB', 'PreProB', 
                                                    'ProB', 'PreB', 'B', #'B1', 'B2', 'B3', 'B4', 'B5', 'B6', 
                                                    'T', 'NK', 'EryP', 'Mono', 'Gr'))) %>% 
```


## B-ALL vs MPAL(B/M)

```{r}
panleucohort <- readRDS("../../../../../../../Other/PanLeuGeneExpr/panLeuTotal-rlog.rds")
panleucohort %>% dim()
```

```{r}
panleucohort_anno <- data.table::fread('../../../../../../../Other/PanLeuGeneExpr/PanLeuCohort.csv') 
panleucohort_anno
```

```{r}
panleucohort_BALL_MPAL <- panleucohort[,colnames(panleucohort) %in% filter(panleucohort_anno, disease %in% c('B-ALL', 'MPAL'))$sample]
panleucohort_BALL_MPAL %>% dim()
```


```{r}
modelweights_BDev <- read_csv('survival_analysis/NMF_Lasso_ModelWeights_BDevOnly.csv')
modelweights_BDev
```

```{r}
panleucohort_rlog_NMFscores <- calculate_NMFscores(panleucohort_BALL_MPAL, modelweights_BDev, scale = T, sampleID = 'sample')
panleucohort_rlog_NMFscores
```


```{r}
panleucohort_rlog_NMFscores %>% left_join(panleucohort_anno) %>% write_csv('panleucohort_BALL_MPAL_rlog_DevState_scores_withAnno.csv')
panleucohort_rlog_NMFscores
```

```{r}
panleucohort_anno %>% filter(disease == 'T-ALL') %>% pull(subtype) %>% table()
```


```{r}
panleucohort_rlog_NMFscores %>% 
  left_join(panleucohort_anno) %>% filter(disease %in% c('T-ALL', 'MPAL')) %>% 
  ggplot(aes(x = reorder(subtype, -Early_Lymphoid), y = Early_Lymphoid)) + 
  geom_boxplot() + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0))
```

```{r}
panleucohort_rlog_NMFscores %>% 
  left_join(panleucohort_anno) %>% filter(disease %in% c('T-ALL', 'MPAL')) %>% 
  ggplot(aes(x = reorder(subtype, -PC1_Early_vs_Committed), y = PC1_Early_vs_Committed)) + 
  geom_boxplot() + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0))
```






## Evaluate on Pharmacotypes 

```{r}
# require gene symbol column to be named "Gene"
rpkm_to_logTPM <- function(dat){
  # convert to TPM
  dat_TPM <- dat %>% 
    gather(-Gene, key = "Sample", value = "RPKM") %>%
    group_by(Sample) %>% 
    mutate(logTPM = log1p(RPKM / sum(RPKM) * 1000000)) %>% 
    select(-RPKM) %>% ungroup() %>% 
    spread(Sample, logTPM)
  
  return(dat_TPM)
}
```


```{r}
pharmacotype_fpkm <- data.table::fread('pharmacotypes/pharmacotyping_ped_rnaseq_fpkm_ALLids_0823.csv') %>% select(-GeneID) %>% dplyr::rename(Gene = GeneName)
pharmacotype_fpkm
```

```{r}
pharmacotype_logTPM <- pharmacotype_fpkm %>% rpkm_to_logTPM()
pharmacotype_logTPM <- pharmacotype_logTPM %>% column_to_rownames('Gene') %>% data.matrix()
pharmacotype_logTPM %>% dim()
```


```{r}
pharmacotype_logTPM_scored <- calculate_NMFscores(pharmacotype_logTPM, modelweights, scale = T, sampleID = 'Patient ID')
pharmacotype_logTPM_scored %>% write_csv('pharmacotypes/ALL_pharmacotypes_logTPM_DevState_scores.csv')
pharmacotype_logTPM_scored
```


```{r}
pharmacotypes <- read_csv('pharmacotypes/ALL_invitro_pharmacotypes.csv')
pharmacotypes
```


```{r}
pharmacotypes_combined <- pharmacotypes %>% inner_join(pharmacotype_logTPM_scored) %>% filter(Immunophenotype == 'B')
pharmacotypes_combined
```

```{r}
CellType_Drug_Corr <- cor(x = select(pharmacotypes_combined, contains('normalized')), 
                          y = select(pharmacotypes_combined, HSC_MPP, Myeloid_Prog, Pre_pDC, Early_Lymphoid, 
                                     Pro_B, Pre_B, Mature_B, T_NK, Monocyte, Erythroid), use = 'pairwise.complete.obs', method = 'spearman')
CellType_Drug_Corr
```

```{r}
CellType_Drug_Corr
```

```{r}
library(corrplot)
CellType_Drug_Corr %>% t() %>% corrplot()
```


## Jae Kim B-ALL Ph+

```{r}
Kim_PhALL_samples <- list.files('subtype_subcluster/Kim2023_Ph_BALL/RNAseq_rawcounts/')
Kim_PhALL_samples
```

```{r}
Kim_PhALL_anno <- read_csv('subtype_subcluster/Kim2023_Ph_BALL/Ph_BALL_ClinicalAnno.csv')
Kim_PhALL_anno
```

```{r}
kim_phALL_counts <- data.table::fread(paste0('subtype_subcluster/Kim2023_Ph_BALL/RNAseq_rawcounts/', Kim_PhALL_samples[1])) 
colnames(kim_phALL_counts) <- c('ENSG', Kim_PhALL_samples[1] %>% str_replace('_count_sub.txt',''))

for(ph_samp in Kim_PhALL_samples[-1]){
  print(ph_samp)
  # Load file 
  temp <- data.table::fread(paste0('subtype_subcluster/Kim2023_Ph_BALL/RNAseq_rawcounts/', ph_samp))
  colnames(temp) <- c('ENSG', ph_samp %>% str_replace('_count_sub.txt','')) 
  # merge
  kim_phALL_counts <- kim_phALL_counts %>% left_join(temp)
}

kim_phALL_counts
```

```{r}
ENSGconvert <- data.table::fread('../../../../../../../CIBERSORT/newDataSets_Jul2020/preprocessing/GRCh38_transcript_lengths.txt')
ENSGconvert <- ENSGconvert %>% select(ENSG = V1, Gene = V2) %>% unique()
ENSGconvert
```

```{r}
kim_phALL_counts <- kim_phALL_counts %>% inner_join(ENSGconvert) %>% select(-ENSG) %>% select(Gene, everything()) %>% 
  group_by(Gene) %>% summarise_all(sum)
kim_phALL_counts
```

```{r}
kim_phALL_counts %>% write_csv('subtype_subcluster/Kim2023_Ph_BALL/Kim2023_Ph_BALL_RNAseq_counts.csv')
```


```{r}
library(DESeq2)

kim_phALL_vst <- DESeqDataSetFromMatrix(kim_phALL_counts %>% column_to_rownames('Gene') %>% data.matrix(), 
                       colData = data.frame('Patient' = colnames(kim_phALL_counts)[-1]) %>% column_to_rownames('Patient'), 
                       design = ~1) %>% vst() %>% assay()
rm(kim_phALL_counts)

kim_phALL_vst[1:10,1:10]
```

```{r}
kim_phALL_vst %>% as.data.frame() %>% rownames_to_column('Gene') %>% write_csv('subtype_subcluster/Kim2023_Ph_BALL/Kim2023_Ph_BALL_RNAseq_vst.csv')
```


```{r}
# calculate NMF scores from vst-normalized data
kim_phALL_vst_NMFscores <- calculate_NMFscores(kim_phALL_vst, modelweights, scale = T, sampleID = 'Sample')
kim_phALL_vst_NMFscores[1:10,1:10]
```

```{r}
kim_phALL_vst_NMFscores %>% write_csv('subtype_subcluster/Kim2023_Ph_BALL/Kim2023_Ph_BALL_LineageNMF_Scored.csv')
```


# compare

```{r}
Kim_PhALL_anno %>% mutate(Sample = ifelse(Manuscript_name %>% str_detect('-R'), paste0(JAMLR, '_nn_M'), paste0(JAMLR, '_nn_P'))) %>% 
  write_csv('subtype_subcluster/Kim2023_Ph_BALL/Kim2023_Ph_BALL_anno_cleaned.csv')
```


```{r}
Kim_PhALL_anno_LineageScored <- Kim_PhALL_anno %>% mutate(Sample = ifelse(Manuscript_name %>% str_detect('-R'), paste0(JAMLR, '_nn_M'), paste0(JAMLR, '_nn_P'))) %>% 
  select(Sample, WBC, Age_at_dx, subtype, Subgroup) %>% 
  inner_join(kim_phALL_vst_NMFscores)

Kim_PhALL_anno_LineageScored
```

```{r}
Kim_PhALL_anno %>% pull(Age_at_dx) %>% summary()
```


```{r, fig.height = 5, fig.width = 12}
Kim_PhALL_anno_LineageScored %>% 
  select(-Sample, -WBC, -Age_at_dx, -subtype) %>% 
  pivot_longer(-Subgroup) %>% 
  ggplot(aes(x = Subgroup, y = value, fill = Subgroup)) + 
  geom_boxplot() + ggbeeswarm::geom_quasirandom() +
  theme_pubr(legend = 'none') + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  facet_wrap(.~name, scales = 'free', ncol = 5) + stat_compare_means()
```









```{r}
CBfractions_vst_NMFscores
```


```{r, fig.height = 5, fig.width = 10}
keep_populations <- c('HSC', 'MPP', 'LMPP', 'CMP', 'GMP', 'MLPII', 'EarlyProB', 'PreProB', 
                      'ProB', 'PreB', 'B')#, 'B2', 'B2_1', 'B2_2',#'IgMminus', 'IgMplus', #'B1', 'B2', 'B3', 'B4', 'B5', 'B6', 
                      #'T', 'NK', 'EryP', 'Mono', 'Gr')
NMFlevels = c('HSC_MPP', 'Myeloid_Prog', 'Pre_pDC', 'Early_Lymphoid', 'Pro_B', 'Pre_B')#, 'Mature_B')

CBfractions_vst_NMFscores %>% 
  pivot_longer(-Sample, names_to = 'NMFsig', values_to = 'Score') %>% 
  mutate(Population = Sample %>% str_replace('.*CB_','')) %>%
  filter(NMFsig %in% NMFlevels, 
         Population %in% keep_populations) %>% 
  mutate(NMFsig = factor(NMFsig, levels = NMFlevels), 
         Population = factor(Population %>% str_replace('EarlyProB', 'CLP') %>% str_replace('MLPII', 'MLP') %>% str_replace('B2.*','B2'), 
                             levels = c('HSC', 'MPP', 'LMPP', 'CMP', 'GMP', 'MLP', 'CLP', 'PreProB', 
                                        'ProB', 'PreB', 'B'))) %>% #, #'B2',#'IgMminus', 'IgMplus',#'B1', 'B2', 'B3', 'B4', 'B5', 'B6', 
                                        #'T', 'NK', 'EryP', 'Mono', 'Gr'))) %>% 
  filter(Population != 'NA') %>% 
  ggplot(aes(x = Population, y = Score, fill = Population)) + 
  geom_boxplot() + geom_jitter() +
  theme_pubr(legend = 'none') + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  facet_wrap(.~NMFsig, scales = 'free', ncol = 3)
  
```
