---
title: "B-ALL Rapid Projections"
output: html_notebook
---

Rapid projections for B-ALL data

```{r}
library(Seurat)
library(symphony)
library(tidyverse)
```

**Bone Marrow Reference Map**

```{r}
BM_projection_path = '../../../../../../AMLhierarchies/scRNA_projection/Complete_Hematopoiesis/BM_V2only/Final_Embedding/Final_projections/ProjectionTools/'
source(paste0(BM_projection_path, 'Symphony_Utils_BMref.R'))
BM_ref <- readRDS(paste0(BM_projection_path, 'BoneMarrow_RefMap_SymphonyRef.rds'))

# Fix uwot path for UMAP projection - point it to the directory with Projection Tools
BM_ref$save_uwot_path <- paste0(BM_projection_path, BM_ref$save_uwot_path)
```

```{r, fig.height = 4.5, fig.width = 10}
ReferenceSeuratObj_BM <- CreateReferenceSeuratObj(BM_ref)
DimPlot(ReferenceSeuratObj_BM, reduction = 'umap', group.by = 'CellType_Annotation_formatted', raster=FALSE, label=TRUE, label.size = 4)
```

**Cross-Ontogeny B-Development Map**

```{r}
Bdev_projection_path = 'BDevelopment_ProjectionTools/'
Bdev_ref <- readRDS(paste0(Bdev_projection_path, 'BDevelopment_RefMap_SymphonyRef.rds'))

# Fix uwot path for UMAP projection - point it to the directory with Projection Tools
Bdev_ref$save_uwot_path <- paste0(Bdev_projection_path, Bdev_ref$save_uwot_path)
```

```{r, fig.height = 5, fig.width = 11}
ReferenceSeuratObj_Bdev <- CreateReferenceSeuratObj(Bdev_ref)
# flip UMAP1 axis
ReferenceSeuratObj_Bdev[['umap']]@cell.embeddings[,1] = -ReferenceSeuratObj_Bdev[['umap']]@cell.embeddings[,1]
DimPlot(ReferenceSeuratObj_Bdev, reduction = 'umap', group.by = 'BDevelopment_CellType_Comprehensive', raster=FALSE, label=TRUE, label.size = 5)
```


### Run Samples

```{r}
library(ggpointdensity)
library(viridis)
library(jcolors)
```

```{r}
map_sample_BM <- function(seurat_obj, batch_key){
  
  # Map cells
  seurat_obj <- mapQuery(
      seurat_obj@assays$RNA@counts, 
      seurat_obj@meta.data,
      BM_ref,
      vars = batch_key, 
      return_type = 'Seurat'
  ) %>% calcMappingError(., reference = BM_ref, MAD.threshold = 2)
  
  # Predict Labels 
  seurat_obj <- knnPredict.Seurat(
    seurat_obj, 
    BM_ref, 
    label_transfer = 'CellType_Annotation', 
    k = 30
  ) 
  
  return(seurat_obj)
}
```

```{r}
library(ggpointdensity)
library(jcolors)

plot_projection <- function(dat, ptx_id, refUMAP = refUMAP, downsample_reference = TRUE, save_folder = 'BALL_projections/Complete_BoneMarrowReference/Figures/'){
  
  if(downsample_reference){
    refUMAP <- refUMAP %>% sample_frac(0.25)
  }
  
  dat <- dat@meta.data %>% rownames_to_column('Cell') %>% 
    left_join(dat@reductions$umap@cell.embeddings %>% data.frame() %>% rownames_to_column('Cell') ) %>% 
    filter(mapping_error_QC == 'Pass') %>% 
    mutate(ref_query = 'query') %>% 
    bind_rows(refUMAP)
  
  ## Get the Background
  background <-  dat %>% filter(ref_query == 'reference') %>% select(-Sample)
  heatpalette <- heat.colors(12)
  
  p <- dat %>% 
    filter(ref_query == 'query') %>% 
    ggplot(aes(x = umap_1, y = umap_2)) +  
    geom_point(data = background, color='#E3E3E3', size=0.01, alpha=0.5) + 
    geom_pointdensity(size=0.05) +
    scale_color_jcolors_contin("pal3", reverse = TRUE, bias = 1.75) +
    geom_density_2d(alpha=0.4, color='black', h = 1.5, size=0.4) + 
    theme_void() + ggtitle(ptx_id) + 
    theme(strip.text.x = element_text(size=18), legend.position='none') 
  
  ggsave(paste0(save_folder, ptx_id, '_BoneMarrowReference_projectedUMAP.pdf'), height = 4, width = 6)
}
```

```{r}
get_composition_BM <- function(BALL, composition_BM){
  # get composition per cell
  sample_composition <- BALL@meta.data %>% rownames_to_column('barcode') %>% 
    select(barcode, Sample, mapping_error_score, mapping_error_QC, 
           CellType_Annotation, CellType_Annotation_prob)
  
  # concatenate lol
  composition_BM <- bind_rows(composition_BM, sample_composition)
  return(composition_BM)
}
``` 


### Now lets iterate through samples and run

```{r}
basepath = 'BALL_rawdata/Expanded/'
matrixpath = '/filtered_feature_bc_matrix/'
BALL_patients <- list.files(basepath)
BALL_patients
```

```{r}
refUMAP <- data.frame(BM_ref$umap$embedding) %>% rename(umap_1 = X1, umap_2 = X2) %>% mutate(ref_query = 'reference')
BALL_composition_BMref <- data.frame()

# Test the first sample
for(pt_samp in BALL_patients){
  print(paste0('Mapping: ', pt_samp))
  start_time <- Sys.time()
  
  # load in as seurat obj
  BALL <- Seurat::Read10X(paste0(basepath, pt_samp, matrixpath)) %>% CreateSeuratObject()
  BALL$Sample <- pt_samp

  # QC >500 genes, >2500 counts, <8% mt counts
  BALL <- Seurat::PercentageFeatureSet(BALL, pattern = '^MT-', col.name = 'pct.mito')
  BALL <- subset(BALL, nFeature_RNA > 500 & nCount_RNA > 2500 & pct.mito < 8)
  print(paste0('Post QC Cells: n = ', dim(BALL)[2]))

  # Map sample onto reference map and label with 30 KNN, no batch correction
  BALL <- map_sample_BM(BALL, NULL)
  # plot projected cells onto UMAP 
  plot_projection(BALL, pt_samp, refUMAP = refUMAP, downsample_reference = TRUE, save_folder = 'BALL_projections/Complete_BoneMarrowReference/Figures/')
  # get celltype composition
  BALL_composition_BMref <- get_composition_BM(BALL, BALL_composition_BMref)
  # save
  BALL %>% saveRDS(paste0("BALL_projections/Complete_BoneMarrowReference/", pt_samp, "_projected.rds"))
  rm(BALL)
  
  end_time <- Sys.time()
  print(end_time - start_time)
  gc()
}

BALL_composition_BMref %>% write_csv('BALL_projections/Complete_BoneMarrowReference/BALL_celltype_annotations_FullReference.csv')
```

# Project onto B Cell Development 

```{r}
map_sample_BDev <- function(seurat_obj, batch_key){
  # Map cells
  seurat_obj <- mapQuery(
      seurat_obj@assays$SymphonyQuery@counts, 
      seurat_obj@meta.data,
      Bdev_ref,
      vars = batch_key, 
      return_type = 'Seurat'
  ) 
  
  # Predict Labels 
  seurat_obj <- knnPredict.Seurat(
    seurat_obj, 
    Bdev_ref, 
    label_transfer = 'BDevelopment_CellType_Comprehensive', 
    k = 30
  ) 
  return(seurat_obj)
}
```


```{r}
library(ggpointdensity)
library(jcolors)

plot_BDev_projection <- function(dat, ptx_id, refUMAP, downsample_reference = TRUE, save_folder = 'BALL_projections/Focused_BDevelopment/Figures/'){
  
  if(downsample_reference){
    set.seed(123)
    refUMAP <- refUMAP %>% sample_frac(0.5)
  }
  
  dat <- dat@meta.data %>% rownames_to_column('Cell') %>% 
    left_join(dat@reductions$umap@cell.embeddings %>% data.frame() %>% rownames_to_column('Cell') ) %>% 
    filter(mapping_error_QC == 'Pass') %>% 
    mutate(ref_query = 'query') %>% 
    bind_rows(refUMAP)
  
  ## Get the Background
  background <-  dat %>% filter(ref_query == 'reference') %>% select(-Sample)
  heatpalette <- heat.colors(12)
  
  p <- dat %>% 
    filter(ref_query == 'query') %>% 
    ggplot(aes(x = -umap_1, y = umap_2)) +  
    geom_point(data = background, color='#E3E3E3', size=0.005, alpha=0.5) + 
    geom_pointdensity(size=0.02) +
    scale_color_jcolors_contin("pal3", reverse = TRUE, bias = 1.75) +
    geom_density_2d(alpha=0.4, color='black', h = 1.5, size=0.3) + 
    theme_void() + ggtitle(ptx_id) + 
    theme(strip.text.x = element_text(size=18), legend.position='none') 
  
  ggsave(paste0(save_folder, ptx_id, '_BdevelopmentReference_projectedUMAP.pdf'), height = 4, width = 6)
}
```


```{r}
get_UMAPcoordinates <- function(BALL, umap_coordinates){
  # get composition per cell
  sample_coordinates <- BALL@meta.data %>% rownames_to_column('barcode') %>%
    select(barcode, Sample) %>% 
    left_join(BALL@reductions$umap@cell.embeddings %>% data.frame() %>% rownames_to_column('barcode') )
  
  # concatenate 
  umap_coordinates <- bind_rows(umap_coordinates, sample_coordinates)
  return(umap_coordinates)
}
``` 


```{r}
get_composition_Bdev <- function(BALL, composition_Bdev){
  # get composition per cell
  sample_composition <- BALL@meta.data %>% rownames_to_column('barcode') %>%
    select(barcode, Sample, BDevelopment_CellType_Comprehensive, BDevelopment_CellType_Comprehensive_prob)
  
  # concatenate 
  composition_Bdev <- bind_rows(composition_Bdev, sample_composition)
  return(composition_Bdev)
}
``` 



### Approach

For each projected sample we will subset the following celltypes: 
  - HSC, MPP-LMPP, LMPP, Early GMP, MLP, MLP-II
  - Pre-pDC, Pre-pDC Cycling, pDC
  - CLP, Pre-ProB, Pro-B VDJ, Pro-B Cycling, Large Pre-B, Small Pre-B, Immature B, Mature B
  
```{r}
BDev_celltypes <- c('HSC', 'HSC/MPP', 'MPP-MyLy', 'MPP-LMPP', 'LMPP', 'Early GMP', 'MLP', 'MLP-II', 'Pre-pDC', 'Pre-pDC Cycling', 'pDC', 
                    'CLP', 'EarlyProB', 'Pre-ProB', 'Pro-B VDJ', 'Pro-B Cycling', 'Large Pre-B', 'Small Pre-B', 'Immature B', 'Mature B')
```
  
```{r}
projectedpath = 'BALL_projections/Complete_BoneMarrowReference/'
#matrixpath = '/filtered_feature_bc_matrix/'
BALL_patients_projected <- list.files(projectedpath, pattern = 'rds')
BALL_patients_projected
```


```{r}
BALL_patients_projected[1]
BALL <- readRDS(paste0(projectedpath, BALL_patients_projected[1])) 
BALL
```

```{r}
BALL_UMAPcoordinates_BMref <- data.frame()
BALL_UMAPcoordinates_BMref <- get_UMAPcoordinates(BALL, BALL_UMAPcoordinates_BMref)
BALL_UMAPcoordinates_BMref
```

### MAKE SURE TO RECORD UMAP COORDINATES from normal and from BDev

```{r}
refUMAP_bdev <- data.frame(Bdev_ref$umap$embedding) %>% rename(umap_1 = X1, umap_2 = X2) %>% mutate(ref_query = 'reference')
BALL_composition_BDev <- data.frame()
BALL_UMAPcoordinates_BMref <- data.frame()
BALL_UMAPcoordinates_BDev <- data.frame()

# Test the first sample
for(pt_samp in BALL_patients_projected){
  print(paste0('Mapping: ', pt_samp))
  start_time <- Sys.time()
  # load in as seurat obj
  BALL <- readRDS(paste0(projectedpath, pt_samp)) 
  # get UMAP coordinates from BM Reference mapping
  BALL_UMAPcoordinates_BMref <- get_UMAPcoordinates(BALL, BALL_UMAPcoordinates_BMref)
  # subset to include B development lineage
  BALL <- subset(BALL, CellType_Annotation %in% BDev_celltypes)
  # Map sample onto reference map and label with 30 KNN, no batch correction
  pt_samp <- pt_samp %>% str_replace('_projected.rds','')
  BALL <- map_sample_BDev(BALL, NULL)
  # plot projected cells onto UMAP 
  plot_BDev_projection(BALL, pt_samp, refUMAP = refUMAP_bdev, downsample_reference = TRUE, save_folder = 'BALL_projections/Focused_BDevelopment/Figures/')
  # get celltype composition
  BALL_composition_BDev <- get_composition_Bdev(BALL, BALL_composition_BDev)
  # get UMAP coordinates from BDevelopment mapping 
  BALL_UMAPcoordinates_BDev <- get_UMAPcoordinates(BALL, BALL_UMAPcoordinates_BDev)
  # save
  BALL %>% saveRDS(paste0("BALL_projections/Focused_BDevelopment/", pt_samp, "_projected_BDevFocus.rds"))
  rm(BALL)
  
  end_time <- Sys.time()
  print(end_time - start_time)
  gc()
}
# save cell annotations
BALL_composition_BDev %>% write_csv('BALL_projections/Focused_BDevelopment/BALL_celltype_annotations_Bdevelopment.csv')
# save UMAP coordinates for FullReference and BDevelopment
BALL_UMAPcoordinates_BMref %>% write_csv('BALL_projections/Complete_BoneMarrowReference/BALL_UMAPcoordinates_FullReference.csv')
BALL_UMAPcoordinates_BDev %>% write_csv('BALL_projections/Focused_BDevelopment/BALL_UMAPcoordinates_Bdevelopment.csv')
```

```{r}
BALL_UMAPcoordinates_BMref %>% write_csv('BALL_projections/Complete_BoneMarrowReference/BALL_UMAPcoordinates_FullReference.csv')
BALL_UMAPcoordinates_BDev %>% write_csv('BALL_projections/Focused_BDevelopment/BALL_UMAPcoordinates_Bdevelopment.csv')
```




### Get Composition


```{r}
BALL_composition_BMref <- read_csv('BALL_projections/Complete_BoneMarrowReference/BALL_celltype_annotations_FullReference.csv')
BALL_composition_BDev <- read_csv('BALL_projections/Focused_BDevelopment/BALL_celltype_annotations_Bdevelopment.csv')
```

```{r}
BALL_UMAPcoordinates_BMref <- read_csv('BALL_projections/Complete_BoneMarrowReference/BALL_UMAPcoordinates_FullReference.csv')
BALL_UMAPcoordinates_BDev <- read_csv('BALL_projections/Focused_BDevelopment/BALL_UMAPcoordinates_Bdevelopment.csv')
```


```{r}
BALL_UMAPcoordinates_BMref %>% dplyr::rename(FullReference_UMAP1 = umap_1, FullReference_UMAP2 = umap_2)
BALL_UMAPcoordinates_BDev %>% mutate(umap_1 = -umap_1) %>% dplyr::rename(BDevelopment_UMAP1 = umap_1, BDevelopment_UMAP2 = umap_2)
```

```{r}
BALL_composition <- 
  BALL_composition_BMref %>% 
  left_join(BALL_composition_BDev) %>%
  left_join( BALL_UMAPcoordinates_BMref %>% dplyr::rename(UMAP1_FullReference = umap_1, UMAP2_FullReference = umap_2) ) %>% 
  left_join( BALL_UMAPcoordinates_BDev %>% mutate(umap_1 = -umap_1) %>% dplyr::rename(UMAP1_BDevelopment = umap_1, UMAP2_BDevelopment = umap_2) ) %>% 
  mutate(CellType_Annotation = CellType_Annotation %>% as.character(), 
         BDevelopment_CellType_Comprehensive = BDevelopment_CellType_Comprehensive %>% as.character()) %>%
  mutate(Final_CellType = ifelse(is.na(BDevelopment_CellType_Comprehensive), CellType_Annotation, BDevelopment_CellType_Comprehensive),
         Final_CellType_prob = ifelse(is.na(BDevelopment_CellType_Comprehensive_prob), CellType_Annotation_prob, BDevelopment_CellType_Comprehensive_prob)) %>%
  select(Cell = barcode, Directory = Sample, Final_CellType, Final_CellType_prob, everything()) 

BALL_composition %>% write_csv('BALL_projections/BALL_89pt_CellType_FinalAssignments.csv')
BALL_composition
```


```{r}
BALL_composition %>% pull(Directory) %>% table() %>% sort()
```


### Methods

Raw counts from filtered BALL cells were projected onto the bone marrow reference map using Symphony (Kang et al). 
Filtering criteria was: 
  - % mito < 8
  - nCount_RNA > 2500
  - nFeature_RNA > 500
  
Cells assigned to be within the B-cell development lineage: 
  - HSC, MPP-LMPP, LMPP, Early GMP, MLP, MLP-II
  - CLP, Pre-ProB, Pro-B VDJ, Pro-B Cycling, Large Pre-B, Small Pre-B, Immature B, Mature B
  - Pre-pDC, Pre-pDC Cycling, pDC

Were subsetted and projected onto the cross-ontogeny map of B-cell development using Symphony to refine cell type classification along the B cell lineage.
















