---
title: "Marker Gene Selection"
output: html_notebook
---

Let's triangulate marker genes for each of the lineages that we find enrichment in. 
We have three sources of information: 

  1) Marker genes by DESeq on pseudobulk samples 
  2) Correlation with abundance across 89 samples

Let's take the Intersection where we take the intersect of loosely filtered results multiple approaches

These genes will be used as a feature space for manual pseudobulk-based regression
  
**N.B. these are lineages derived from consensus clustering on NMF weights**
  **This dimensionality reduction approach actually lead to cleaner clustering of celltypes than just clustering celltypes by correlation**
  

## Setup 

```{r}
library(tidyverse)
library(ggpubr)
```

Code from AUCell for adaptive thresholding 

```{r}
library(AUCell)

### Helper function for AUC thresholding from AUCell
get_Threshold <- function (auc, gSetName, plotHist = TRUE, smallestPopPercent = 0.25, 
          densAdjust = 2, thrP = 0.01, nBreaks = 100) 
{
  aucRow <- t(as.matrix(auc))
  #gSetName <- rownames(aucRow)[1]
  nCells <- length(auc)
  skipGlobal <- TRUE
  skipRed <- FALSE
  skipSmallDens <- FALSE
  commentMsg <- ""
  aucThrs <- c()
  notPopPercent <- 1 - smallestPopPercent
  if (sum(auc == 0) > (nCells * notPopPercent)) {
    skipGlobal <- FALSE
    commentMsg <- paste(commentMsg, round((sum(auc == 0)/nCells) * 
                                            100), "% (more than ", notPopPercent, "%) of AUC are zero. ", 
                        sep = "")
  }
  meanAUC <- mean(auc)
  sdAUC <- sd(auc)
  maybeNormalDistr <- !suppressWarnings(ks.test(auc, rnorm(max(100, 
                                                               length(auc)), mean = meanAUC, sd = sdAUC), alternative = "less")$p.value < 
                                          0.01)
  if (maybeNormalDistr) {
    commentMsg <- paste0(commentMsg, "The AUC might follow a normal distribution (random gene-set?). ")
    skipGlobal <- FALSE
    aucThrs["outlierOfGlobal"] <- qnorm(1 - (thrP/nCells), 
                                        mean = meanAUC, sd = sdAUC)
  }
  histogram <- hist(c(0, auc/max(auc)), breaks = 100, plot = FALSE)$count
  if ((sum(histogram[1:5])/sum(histogram)) >= notPopPercent * 
      0.75) {
    skipGlobal <- FALSE
    skipRed <- TRUE
    skipSmallDens <- TRUE
  }
  if ((sum(histogram[1:10])/sum(histogram)) >= notPopPercent * 
      0.5) {
    skipSmallDens <- TRUE
    skipGlobal <- FALSE
    aucThrs["tenPercentOfMax"] <- max(auc) * 0.1
  }
  densCurve <- density(auc, adjust = densAdjust, cut = 0)
  maximumsDens <- NULL
  inflPoints <- diff(sign(diff(densCurve$y)))
  maximumsDens <- which(inflPoints == -2)
  globalMax <- maximumsDens[which.max(densCurve$y[maximumsDens])]
  minimumDens <- which(inflPoints == 2)
  smallMin <- NULL
  if (!skipSmallDens) 
    smallMin <- data.table::last(minimumDens[which(minimumDens < 
                                                     globalMax)])
  minimumDens <- c(smallMin, minimumDens[which(minimumDens > 
                                                 globalMax)])
  densTrh <- NULL
  if (length(minimumDens) > 0) {
    densTrh <- densCurve$x[min(minimumDens)]
    if (length(maximumsDens) > 0) {
      nextMaxs <- maximumsDens[which(densCurve$x[maximumsDens] > 
                                       densTrh)]
      if ((max(densCurve$y[nextMaxs])/max(densCurve$y)) < 
          0.05) {
        densTrh <- NULL
      }
    }
  }
  auc <- sort(auc)
  distrs <- list()
  distrs[["Global_k1"]] <- list(mu = c(meanAUC, NA), sigma = c(sdAUC, 
                                                               NA), x = auc)
  if ("mixtools" %in% rownames(installed.packages())) {
    na <- capture.output(distrs[["k2"]] <- tryCatch(mixtools::normalmixEM(auc, 
                                                                          fast = FALSE, k = 2, verb = FALSE), error = function(e) {
                                                                            return(NULL)
                                                                          }))
    na <- capture.output(distrs[["k3"]] <- tryCatch(mixtools::normalmixEM(auc, 
                                                                          fast = FALSE, k = 3, verb = FALSE), error = function(e) {
                                                                            return(NULL)
                                                                          }))
    if (is.null(distrs[["k2"]]) && is.null(distrs[["k3"]])) {
      if (sum(auc == 0) < (nCells * notPopPercent * 0.5)) 
        skipGlobal <- FALSE
    }
    if (!is.null(distrs[["k2"]])) {
      compL <- which.min(distrs[["k2"]][["mu"]])
      compR <- which.max(distrs[["k2"]][["mu"]])
      height1 <- 0.4/distrs[["k2"]][["sigma"]][compL] * 
        distrs[["k2"]][["lambda"]][compL]
      height2 <- 0.4/distrs[["k2"]][["sigma"]][compR] * 
        distrs[["k2"]][["lambda"]][compR]
      taller <- height1 < height2
      globalInclInFirst <- (distrs[["Global_k1"]]$mu[1] < 
                              (distrs[["k2"]][["mu"]][compL] + (1.5 * distrs[["k2"]][["sigma"]][compL])))
      includedInGlobal <- ((distrs[["k2"]][["mu"]][compL] > 
                              (distrs[["Global_k1"]]$mu[1] - distrs[["Global_k1"]]$sigma[1])) && 
                             (distrs[["k2"]][["mu"]][compR] < (distrs[["Global_k1"]]$mu[1] + 
                                                                 distrs[["Global_k1"]]$sigma[1])))
      if (taller || (globalInclInFirst && includedInGlobal)) {
        skipGlobal <- FALSE
        if (globalInclInFirst && includedInGlobal) 
          commentMsg <- paste(commentMsg, "The global distribution overlaps the partial distributions. ")
        if (taller && !includedInGlobal) 
          commentMsg <- paste(commentMsg, "The right distribution is taller. ")
      }
    }
  }
  else {
    warning("Package 'mixtools' is not available to calculate the sub-distributions.")
  }
  glProb <- 1 - (thrP/nCells + smallestPopPercent)
  aucThrs["Global_k1"] <- qnorm(glProb, mean = distrs[["Global_k1"]][["mu"]][1], 
                                sd = distrs[["Global_k1"]][["sigma"]][1])
  if (!is.null(distrs[["k2"]])) {
    k2_L <- which.min(distrs[["k2"]][["mu"]])
    aucThrs["L_k2"] <- qnorm(1 - (thrP/nCells), mean = distrs[["k2"]][["mu"]][k2_L], 
                             sd = distrs[["k2"]][["sigma"]][k2_L])
  }
  if (!is.null(distrs[["k3"]])) {
    k3_R <- which.max(distrs[["k3"]][["mu"]])
    k3_R_threshold <- qnorm(thrP, mean = distrs[["k3"]][["mu"]][k3_R], 
                            sd = distrs[["k3"]][["sigma"]][k3_R])
    if (k3_R_threshold > 0) 
      aucThrs["R_k3"] <- k3_R_threshold
  }
  if (!is.null(densTrh)) {
    aucThrs["minimumDens"] <- densTrh
  }
  aucThr <- aucThrs
  if (skipGlobal) 
    aucThr <- aucThrs[which(!names(aucThrs) %in% "Global_k1")]
  if (skipRed) 
    aucThr <- aucThrs[which(!names(aucThrs) %in% "L_k2")]
  aucThr <- aucThr[which.max(aucThr)]
  if ((length(aucThr) > 0) && (names(aucThr) == "minimumDens")) {
    maximumsDens <- maximumsDens[which(densCurve$y[maximumsDens] > 
                                         1)]
    if (length(maximumsDens) > 2) {
      tmp <- cbind(minimumDens[seq_len(length(maximumsDens) - 
                                         1)], maximumsDens[-1])
      FCs <- densCurve$y[tmp[, 2]]/densCurve$y[tmp[, 1]]
      if (any(FCs > 1.5)) 
        warning(gSetName, ":\tCheck the AUC histogram. ", 
                "'minimumDens' was selected as the best threshold, ", 
                "but there might be several distributions in the AUC.")
    }
  }
  if ("minimumDens" %in% names(aucThrs)) 
    aucThr <- aucThrs["minimumDens"]
  if (length(aucThr) == 0) 
    aucThr <- aucThrs[which.max(aucThrs)]
  if (length(aucThr) == 0) 
    aucThr <- 1
  if (length(aucThr) > 1) 
    aucThr <- unlist(aucThr[which.max(aucThr)])
  if (plotHist) {
    histInfo <- AUCell_plotHist(aucRow, aucThr = aucThr, 
                                nBreaks = nBreaks)
    histMax <- max(histInfo[[gSetName]]$counts)
    densCurve$y <- densCurve$y * (histMax/max(densCurve$y))
    thisLwd <- ifelse((aucThrs["minimumDens"] == aucThr) && 
                        (!is.null(aucThr) && !is.null(aucThrs["minimumDens"])), 
                      3, 1)
    lines(densCurve, lty = 1, lwd = thisLwd, col = "blue")
    if (!is.null(minimumDens)) 
      points(densCurve$x[minimumDens], densCurve$y[minimumDens], 
             pch = 16, col = "darkblue")
    scalFact <- 1
    aucDistr <- dnorm(distrs[["Global_k1"]][["x"]], mean = distrs[["Global_k1"]][["mu"]][1], 
                      sd = distrs[["Global_k1"]][["sigma"]][1])
    scalFact <- (histMax/max(aucDistr)) * 0.95
    thisLwd <- ifelse(aucThrs["Global_k1"] == aucThr, 3, 
                      1)
    lines(distrs[["Global_k1"]][["x"]], scalFact * aucDistr, 
          col = "darkgrey", lwd = thisLwd, lty = 2)
    if (!is.null(distrs[["k2"]])) {
      aucDistr <- dnorm(distrs[["k2"]][["x"]], mean = distrs[["k2"]][["mu"]][k2_L], 
                        sd = distrs[["k2"]][["sigma"]][k2_L])
      scalFact <- (histMax/max(aucDistr)) * 0.95
      thisLwd <- ifelse(aucThrs["k2"] == aucThr, 3, 1)
      lines(distrs[["k2"]][["x"]], scalFact * aucDistr, 
            col = "red", lwd = thisLwd, lty = 2)
      rect(distrs[["k2"]][["mu"]][k2_L] - distrs[["k2"]][["sigma"]][k2_L], 
           histMax - (histMax * 0.02), distrs[["k2"]][["mu"]][k2_L] + 
             distrs[["k2"]][["sigma"]][k2_L], histMax, col = "#70000030", 
           border = "#00009000")
    }
    if ((!is.null(distrs[["k3"]])) && ("R_k3" %in% names(aucThrs))) {
      k3_L <- which.min(distrs[["k3"]][["mu"]])
      aucDistr2 <- dnorm(distrs[["k3"]][["x"]], mean = distrs[["k3"]][["mu"]][k3_R], 
                         sd = distrs[["k3"]][["sigma"]][k3_R])
      scalFact2 <- scalFact * (distrs[["k3"]][["lambda"]][k3_R]/distrs[["k3"]][["lambda"]][k3_L])
      thisLwd <- ifelse(aucThrs["k3"] == aucThr, 3, 1)
      lines(distrs[["k3"]][["x"]], scalFact2 * aucDistr2, 
            col = "magenta", lwd = thisLwd, lty = 2)
      rect(distrs[["k3"]][["mu"]][k3_R] - distrs[["k3"]][["sigma"]][k3_R], 
           histMax - (histMax * 0.02), distrs[["k3"]][["mu"]][k3_R] + 
             distrs[["k3"]][["sigma"]][k3_R], histMax, col = "#80808030", 
           border = "#80808030")
    }
    aucThrs <- aucThrs[!is.na(aucThrs)]
    if (length(aucThrs) > 0) {
      pars <- list()
      pars[["Global_k1"]] <- c(col1 = "#909090", col2 = "black", 
                               pos = 0.9)
      pars[["L_k2"]] <- c(col1 = "red", col2 = "darkred", 
                          pos = 0.8)
      pars[["R_k3"]] <- c(col1 = "magenta", col2 = "magenta", 
                          pos = 0.6)
      pars[["minimumDens"]] <- c(col1 = "blue", col2 = "darkblue", 
                                 pos = 0.4)
      pars[["tenPercentOfMax"]] <- c(col1 = "darkgreen", 
                                     col2 = "darkgreen", pos = 0.9)
      pars[["outlierOfGlobal"]] <- c(col1 = "darkgreen", 
                                     col2 = "darkgreen", pos = 0.9)
      for (thr in names(aucThrs)) {
        thisLwd <- ifelse(aucThrs[thr] == aucThr, 5, 
                          2)
        thisLty <- ifelse(aucThrs[thr] == aucThr, 1, 
                          3)
        abline(v = aucThrs[thr], col = pars[[thr]][1], 
               lwd = thisLwd, lty = thisLty)
        xPos <- aucThrs[thr] * 1.01
        if (aucThrs[thr] > (max(auc) * 0.8)) 
          xPos <- 0
        if (aucThrs[thr] == aucThr) 
          text(xPos, histMax * as.numeric(pars[[thr]][3]), 
               pos = 4, col = pars[[thr]][2], cex = 0.8, 
               paste("AUC > ", signif(aucThrs[thr], 2), 
                     "\n(", sum(auc > aucThrs[thr]), " cells)", 
                     sep = ""))
      }
    }
  }
  return(list(selected = aucThr, thresholds = cbind(threshold = aucThrs, 
                                                    nCells = sapply(aucThrs, function(x) sum(auc > x))), 
              comment = commentMsg))
}

```


## DE Genes

```{r}
library(Seurat)
library(tidyverse)
```

```{r}
Lineage_DE <- read_csv("BALL_DEresults_NMF_Lineage.csv")
Lineage_DE
```

```{r}
Lineage_DE %>% 
  filter(padj < 0.01) %>% 
  pull(Lineage) %>% table() %>% sort(decreasing = T)
```


## For NMF score Quantification

Pearson correlation VST NMF stringent
  Intersect on Pseudobulk DE FDR 0.05
  
Or LASSO regression
  Pearson correlation stringent VST NMF + Pseudobulk DE stringent
  
**Set adaptive thresholds**  
Ths is the pearson correlation between VST-normalized gene expression and the score for each NMF lineage

```{r}
NMF_corr <- data.table::fread('NMF_gene_corr.csv') %>% select(-V1)
NMF_corr
```

```{r}
NMF_corr %>% 
  filter(qvalue < 0.05, pearson > 0) %>% 
  pull(NMF) %>% table() %>% sort(decreasing = TRUE)
```

##### Set Positive Correlation Thresholds

```{r}
NMF_corr_thresholds = data.frame()

for(lin in unique(NMF_corr$NMF)){
  thresholds = get_Threshold(NMF_corr %>% filter(NMF == lin, qvalue < 0.05, pearson > 0) %>% pull(pearson), lin)$thresholds
  NMF_corr_thresholds = 
    bind_rows(
      NMF_corr_thresholds,
      data.frame(
        'NMF' = lin,
        'K1_threshold' = thresholds['Global_k1','threshold'],
        'K2_threshold' = thresholds['L_k2','threshold']
    ))
}
```


Visualize thresholding within all positive correlations

```{r, fig.height=3, fig.width=12}
NMF_corr_thresholds %>% 
  left_join(NMF_corr) %>% filter(pearson > 0) %>% 
  mutate(NMF = factor(NMF, levels = c('NMF6', 'NMF8', 'NMF2', 'NMF1', 'NMF3', 'NMF9', 'NMF4',
                                      'NMF5', 'NMF10', 'NMF7'))) %>% 
  mutate(threshold = ifelse(pearson > K1_threshold, 'pass', 'fail')) %>% 
  ggplot(aes(x = pearson, fill = threshold)) + 
  geom_histogram(bins=100) + theme_pubr(legend = 'top') + 
  scale_fill_brewer(palette = 'Dark2', direction = -1) + 
  facet_wrap(.~NMF, scale = 'free', ncol=5) + 
  geom_vline(aes(xintercept = K1_threshold), lty=2)

```

Visualize thresholding within FDR < 0.05 correlations

```{r, fig.height=3, fig.width=12}
NMF_corr_thresholds %>% 
  left_join(NMF_corr) %>% filter(qvalue < 0.05, pearson > 0) %>% 
  mutate(NMF = factor(NMF, levels = c('NMF6', 'NMF8', 'NMF2', 'NMF1', 'NMF3', 'NMF9', 'NMF4',
                                      'NMF5', 'NMF10', 'NMF7'))) %>% 
  mutate(threshold = ifelse(pearson > K1_threshold, 'pass', 'fail')) %>% 
  ggplot(aes(x = pearson, fill = threshold)) + 
  geom_histogram(bins=100) + theme_pubr(legend = 'top') + 
  scale_fill_brewer(palette = 'Dark2', direction = -1) + 
  facet_wrap(.~NMF, scale = 'free', ncol=5) + 
  geom_vline(aes(xintercept = K1_threshold), lty=2)

```

```{r}
NMF_corr_pos <- NMF_corr_thresholds %>% 
  left_join(NMF_corr) %>% 
  mutate(threshold = ifelse(pearson > K1_threshold, 'pass', 'fail')) 

NMF_corr_pos %>% 
  filter(qvalue < 0.05, pearson > 0) %>% 
  filter(threshold == 'pass') %>% 
  pull(NMF) %>% table() %>% sort(decreasing = T)
```

### Negative thresholds

```{r}
NMF_corr_neg_thresholds = data.frame()

for(lin in unique(NMF_corr$NMF)){
  thresholds = get_Threshold(NMF_corr %>% filter(NMF == lin, qvalue < 0.05, pearson < 0) %>% mutate(pearson = -pearson) %>% pull(pearson), lin)$thresholds
  NMF_corr_neg_thresholds = 
    bind_rows(
      NMF_corr_neg_thresholds,
      data.frame(
        'NMF' = lin,
        'K1_threshold' = thresholds['Global_k1','threshold'],
        'K2_threshold' = thresholds['L_k2','threshold']
    ))
}
```


Visualize thresholding within all negative correlations

```{r, fig.height=3, fig.width=12}
NMF_corr_neg_thresholds %>% 
  left_join(NMF_corr) %>% filter(pearson < 0) %>% 
  mutate(negative_pearson = -pearson) %>% 
  mutate(NMF = factor(NMF, levels = c('NMF6', 'NMF8', 'NMF2', 'NMF1', 'NMF3', 'NMF9', 'NMF4',
                                      'NMF5', 'NMF11', 'NMF10', 'NMF7'))) %>% 
  mutate(threshold = ifelse(negative_pearson > K1_threshold, 'pass', 'fail')) %>% 
  ggplot(aes(x = negative_pearson, fill = threshold)) + 
  geom_histogram(bins=100) + theme_pubr(legend = 'top') + 
  scale_fill_brewer(palette = 'Dark2', direction = -1) + 
  facet_wrap(.~NMF, scale = 'free', ncol=6) + 
  geom_vline(aes(xintercept = K1_threshold), lty=2)

NMF_corr_neg_thresholds
```

Visualize thresholding within FDR < 0.05 correlations

```{r, fig.height=3, fig.width=12}
NMF_corr_neg_thresholds %>% 
  left_join(NMF_corr) %>% filter(qvalue < 0.05, pearson < 0) %>% 
  mutate(negative_pearson = -pearson) %>% 
  mutate(NMF = factor(NMF, levels = c('NMF6', 'NMF8', 'NMF2', 'NMF1', 'NMF3', 'NMF9', 'NMF4',
                                      'NMF5', 'NMF11', 'NMF10', 'NMF7'))) %>% 
  mutate(threshold = ifelse(negative_pearson > K1_threshold, 'pass', 'fail')) %>% 
  ggplot(aes(x = negative_pearson, fill = threshold)) + 
  geom_histogram(bins=100) + theme_pubr(legend = 'top') + 
  scale_fill_brewer(palette = 'Dark2', direction = -1) + 
  facet_wrap(.~NMF, scale = 'free', ncol=6) + 
  geom_vline(aes(xintercept = K1_threshold), lty=2)

NMF_corr_neg_thresholds
```

```{r}
NMF_corr_neg <- NMF_corr_neg_thresholds %>% 
  left_join(NMF_corr) %>% 
  mutate(threshold = ifelse(pearson < -K1_threshold, 'pass', 'fail')) 

NMF_corr_neg %>% 
  filter(qvalue < 0.05, pearson < 0) %>% 
  filter(threshold == 'pass') %>% 
  pull(NMF) %>% table() %>% sort(decreasing = T)
```

```{r}
NMF_corr_neg %>% 
  filter(qvalue < 0.05, pearson < 0) 
```

```{r}
NMF_corr_thresholding <- NMF_corr_pos %>% select(NMF, Gene, pearson, pvalue, qvalue, pos_K1_threshold = K1_threshold, pos_threshold = threshold) %>% 
  left_join(NMF_corr_neg %>% select(NMF, Gene, neg_K1_threshold = K1_threshold, neg_threshold = threshold)) %>% 
  mutate(neg_K1_threshold = -neg_K1_threshold, threshold = ifelse(pos_threshold == 'pass' | neg_threshold == 'pass', 'pass', 'fail')) %>% 
  select(NMF, Gene, pearson, pvalue, qvalue, pos_K1_threshold, neg_K1_threshold, threshold)

NMF_corr_thresholding
```


```{r}
#NMF_corr_thresholding %>% write_csv("NMF_GeneCorr_Thresholding.csv")
```
































